Verification of Stencil Computations through Spatial Specifications

### 2016-07-07 paper.tex
tweaks and spelling

\begin{abstract}
  Verifying the correctness of numerical computations relies on first
  giving a specification of their behaviour. In many cases, a
  numerical computation is derived from an underlying mathematical
  model, but this is often difficult to use as a specification as it
  is highly decoupled from the implementation. We have been studying
  specification techniques in the middle ground between low-level and
  high-level specifications for numerical code. In this paper, we
  focus on \emph{stencil computations}---a common idiom in numerical
  code, but one which is error-prone due to fine-grained indexing
  errors. We hypothesise that in practise, stencil computations tend
  to have a regular shape which is amenable to static analysis and
  simple specification. We describe a abstract spatial specification
  language for stencil computations that can be embedded as
  annotations in source code, along with inference, checking, and
  specification synthesis procedures. We evaluate our
  language against a corpus of numerical Fortran code, and show that
  the vast majority of stencil computations indeed have a simple,
  regular, static shape. Around 10\% of the stencils found evoke a
  programming pattern for which indexing errors are plausible. We
  investigate two stencils in detail and find that we are able to
  detect the vast majority of potential errors. This paper details our
  stencil specification language, its static semantics, inference,
  checking, and synthesis procedures, our implementation, and
  our evaluation study.
\end{abstract}

@@

\item we provide an implementation of our approach as an extension to
  CamFort, an open-source program analysis tool for Fortran programs.
  We use this for a quantiative evaluation of our approach over a corpus of
  scientific applications, totalling roughly one million lines of
  code;

@@

In recent years, there has been an increased awareness
of the need for formal verification techniques in
science~\cite{post2005computational,oberkampf2010verification,orchard2014computational}.
Our specification language is an approach in this direction, providing
a system of lighweight specification for numerical code.
The aim is to not only provide a way of exposing errors, but also
for specifying program behaviour to future-proof the code and ease
its maintenance. In order to increase the likelihood of impact, we aim
to integrate with existing practises smoothly, hence our
specifications are integrated as comments and our tool can synthesis
specifications for the user.

@@

\begin{definition}[Neighbourhood index]
  For an array subscript $v(\bar{e})$, we say that $e \in \bar{e}$
  is a \emph{neighbourhood index} if it is of the form
  $e \equiv i$, $e \equiv i + a$ or $e \equiv i - a$ where $a$ is an
  integer constant. That is, a neighbourhood index is a
  relative index defined as constant translations of base induction
  variables. (The relation $\equiv$ hre identifies terms up-to commutativity
  of $+$ and the inverse relation of $+$ and $-$ \eg{},
  $(-b) + i \equiv i - b$).  We classify neighbourhood indices using
  the predicate \neigh{}. %An array subscript $v(\bar{e})$ where every
% $e \in \bar{e}$ is a neighbourhood index is called a
% \emph{neighbourhood subscript}.
\label{def:neighbour}
\end{definition}

@@

Figure~\ref{fig:equations} lists the equational theory for
specifications which is defined by the binary relation $\equiv$
between region terms. The equations use abbreviations
\term{refl}, \term{fwd}, \term{bwd}, \term{cen}, and \irreflS{}
for \term{reflexive}, \term{forward}, \term{backward},
\term{centered}, and \irrefl{} respectively. Futhermore, the abbreviations include syntactic sugar
for quantifying over the abseence or presence of the \irrefl{}
attribute on region constants as a boolean. For example,
the syntactic sugar for \term{fwd} is defined:
%%
\begin{align*}
\hspace{-0.25em}
& \stenFwdSR{n}{d}{r}
\; := \; \\[-0.4em]
& \qquad \begin{cases}
\stenFwd{n}{d}{, \irrefl} & \;\;\textit{if} \;\, \neg \, r \\
\stenFwd{n}{d}{}  & \;\;\textit{if} \;\,  r
\end{cases} \\
\textit{or} \;\;
& \stenFwdS{n}{d} := \stenFwd{n}{d}{}
\end{align*}
%
and similarly for \term{bwd} and \term{cen}.

@@

\paragraph{Overlapping regions with +}
The first nine
equations (with labels of the form $\ldots \texttt{+} \ldots$)
explain the notion of \emph{region overlapping} with sum
\term{+}. In the first six, $r$ and $s$ range
over booleans denoting the absence (true) or presence (false)
of \irrefl{}. These are combined via disjunction. For example,
the rule $(\textsc{B} \, \term{+} \,
\textsc{F})$ explains that the sum of \term{backward}
and \term{forward} regions in the same dimension and the same
depth is equivalent to a \term{centered} region of that depth.
If the \term{forward} part is irrefleixve ($r =
\textit{false}$) but \term{backward} is not ($s =
\textit{true}$) then their sum is equal to a \term{centered}
region with no \irreflS{} modifier by the disjunction ($r \vee s =
\textit{true}$), \ie{}:
%%
\begin{align*}
& \stenCenS{n}{d}  \\[-0.2em]
\equiv \;\; & \stenFwdSR{n}{d}{\irreflS} \; \texttt{+} \; \stenBwdS{n}{d}
\end{align*}
%%
That is, the \term{fwd} region is irreflexive describing code which is missing
a zero-offset neighbourhood index (\eg{}, missing $\texttt{a(i)}$), but the addition of
the \term{bwd} has no irreflexive attribute and thus provides the
missing zero offset index.

@@

Figure~\ref{fig:inequations} defines a notion of sub-specifications
via the relation $<:^r$ on \textit{region}
syntax and then $<:$ on \textit{spec} syntax inductively.
Both relations are reflexive and transitive (we omit
these equations) and congruent
with respect to \term{+}, \term{*}, \term{atLeast}, \term{atMost},
and \term{readOnce}. The (\trule{\textsc{eq}}) rule connects region
equations with inequations. The (\trule{\textsc{over}}) rule explains
the notion of spatial over-approximation via overlapping regions.
The (\trule{\textsc{rep}}) allows the modifier that
enforces that every index is read just once to be dropped
as an approxiamtion; (\trule{\textsc{bound}}) gives
a connection between the \term{atLeast} and \term{atMost}
approximation modifiers.

@@

Indexing schemes give an abstract
representation of index expressions in the source language.
Our model represents the meaning of stencil specifications
as indexing schemes, which explain the range of possible
indexing behaviours allowed under the specification.
The relationship between indexing schemes and source language
terms is given precisely by the partial function, \textsf{schematic}, which
maps syntactic array-index terms to indexing schemes. This
operation is used by the inference
and the checking procedures.
\end{definition}
%
\begin{definition}%[Schematic]
\label{def:schematic}
The partial function $\textsf{schematic}$ maps
source indexing terms $(e_1, .., e_n)$
to $n$-dimensional index schemes, defined:
%
\begin{align*}
\mathsf{schematic}(e_1, \ldots, e_n) & =
[\downarrow\!e_1 \ldots \downarrow\!e_n ]
\quad \textit{iff} \;\; \forall i . \downarrow\!e_i \neq \bot
\end{align*}
%
The $\downarrow$ operator is partial
 $\downarrow : \textit{expr} \rightarrow (\domainVal + \bot)$
mapping individal indexing expressions to 
an indexing scheme component, defined:
%%
\begin{align*}
\downarrow\!e
 & =  \begin{cases}
a & e \equiv i + a \; \wedge \; \mathsf{neigh}(i + a) \\
-a & e \equiv i - a \; \wedge \; \mathsf{neigh}(i - a) \\
\infty & \textit{$\mathsf{IV}(e) = \emptyset$} \\
\bot   & \textit{otherwise}
\end{cases}
\end{align*}
%%
where $\mathsf{IV}(e)$ is the set of free induction variables
in an expression, and $\neigh{}$ classifies expressions which
are constant offsets from an induction variable
(Defn.~\ref{def:neighbour}). In the third case,
if the expression is not a neighbourhood index but is not
a relative index either (\ie{}, not defined in terms
of an induction variable) then it is mapped to $\infty$, otherwise
$\downarrow\!e$ is undefined (returns $\bot$). The $\mathsf{schematic}$ function
is undefined if $\downarrow$ of any of its components $e_i$ is
undefined.

@@

The model is given by the interpretation function $\interp{-}_n$
(where $n$ is the maximum dimensionality of the specification being
modelled) mapping closed\footnote{That is, we assume that there are no
  occurences of \textit{rvar} in a specification being modelled.  Any
  \emph{open} specification containing region variables can be made
  closed by straightforward syntactic subtitution with a (closed)
  \textit{region}.} specifications to sets of
$n$-dimensional indexing schemes. The interperation is overloaded on
\emph{regions} in Figure~\ref{fig:region-model} and on the top-level
of a specification \textit{spec} in Figure~\ref{fig:spatial-model}.

@@

\begin{example}
For a $2$-dimensional stencil computation, then
\begin{align*}
& \interp{\stenCenS{2}{1}}_2 \\[-0.1em]
& \qquad = \; \{i\singleEntry{1}{2} \,|\, i \in \{\text{-}2,..., 2\}\!\setminus\!0 \}
  \cup  \{ \zeroEntry{1}{2} \,|\, \textit{true} \} \\[-0.2em]
& \qquad = \; \{-2\singleEntry{1}{2}, \, -\singleEntry{1}{2}, \,
  \singleEntry{1}{2}, \, 2\singleEntry{1}{2}\} \cup
  \{\zeroEntry{1}{2}\} \\[-0.2em]
& \qquad = \; \{\vtwoh{-2}{\infty}, \vtwoh{-1}{\infty}, \vtwoh{0}{\infty},
\vtwoh{1}{\infty}, \vtwoh{2}{\infty}\}
\end{align*}
Note the absorbing behaviour of $\infty$ with respect to
multiplication.
\end{example}
\noindent
The last two equations of Fig.~\ref{fig:region-model} give the
models of $\term{+}$ and $\term{*}$. For $\term{+}$ this is the 
straightforward union of the models of subterms. The model of
$\term{*}$ takes the tensor $\otimes$ of subterm models. The tensor is
non-trivial; we give some informal intutition first.

@@

The intuition behind $\otimes$ is that it takes
all possible pairs of indexing schemes, treating them as the
lower and upper bounds of an $n$-dimensional rectangle from which the
remaining vertices of the rectangle are generated (like a bounding
box) \eg{}:
%%
\begin{align*}
\{\vtwohs{1}{2}, \vtwohs{5}{6}\,\} \otimes \{\vtwohs{3}{4}\} =
\{\vtwohs{1}{2}, \vtwohs{3}{2}, \vtwohs{1}{4}, \vtwohs{3}{4}, 
   \vtwohs{3}{6}, \vtwohs{5}{4}, \vtwohs{5}{6}\,\}
\end{align*}
%%
The tensor takes a Cartesian product of the two models,
giving all pairs of schemes which are combined using the
\emph{pairwise permutation} written $\bowtie$ (defined later).
Pairwise permutation generates all possible interleavings of two vectors
(\ie{}, all possible non-deterministic choices between $u_i$ and
$v_i$ in each dimension). Some care is then taken over
how this interacts with $\infty$: pairwise permtuation
corresponds to non-determinstic choice, in each dimension $i$,
between two values $u_i$ and $v_i$, but if either
is $\infty$ then the resulting pairwise permutation is
filtered out if the dimension $i$ has \emph{any other scheme
in the model where $u_i \neq \infty$}, \eg{}:
\begin{align*}
\{\vtwoh{1}{\infty}\} \otimes \{\vtwoh{3}{4}\} =
\{\vtwoh{1}{4}, \vtwoh{3}{4}\}
\end{align*}
%
where $\vtwoh{1}{\infty}$ and $\vtwoh{3}{\infty}$ have been filtered.
%With respect to the notion of unification, the
%
%
% combines non-$\infty$ constraints
%of the unifiers in different dimensions into less accepting unifiers.
%This means an access pattern generalises with this scheme would require
%both models to be satisfied. Use of $dims$ gives the dimensions that
%constitutent models talk about and is used in the product to eliminate
%general unifier appearing in these dimensions. Allowing general unifier
%in these dimensions would prevent both models to be satisfied
%simultaneously.
%  u in M * N   u < w
% mplies that
%  u_1 in M  u1 < w
%  u_2 in M  u2 < w
%
The notion of which dimensions $i$ have that \emph{any scheme in the model
has $u_i \neq \infty$} is represented by the \emph{constrained
dimensions} intermediate function:
%
\begin{definition}%[Constrained dimensions]
For an $n$-dimensional model $M$, its \emph{constrained dimensions}
is the set of dimension identifiers for which at least one scheme $u \in M$ has
a non-$\infty$ value in that dimension. That is:
%
\begin{align*}
\effdims{M}{n} =
\bigcup_{u \in M} \{i \mid i \in \{1,\ldots,n\} \, \wedge \, u_i \neq
  \infty\}
\end{align*}
For example, $\effdims{\{\vtwoh{0}{\infty},\vtwoh{\infty}{\infty}\}}{2}
 = \{1\}$.
\end{definition}

@@

The \emph{pairwise permutation} $u \bowtie v$ 
builds a $2^n \times n$ matrix 
of all possible $n$-vectors generated by
non-deterministically picking for each $j^{th}$ entry either 
$u_j$ or $v_j$ \eg{}
%
\begin{equation*}
\vthreeh{0}{1}{2} \bowtie \vthreeh{3}{4}{5} =
\left[
\setlength{\arraycolsep}{0.42em}
\begin{array}{cccccccc}
0 & 0 & 0 & 0 & 3 & 3 & 3 & 3 \\
1 & 1 & 4 & 4 & 1 & 1 & 4 & 4 \\
2 & 5 & 2 & 5 & 2 & 5 & 2 & 5
\end{array}
\right]
% \\
%\setlength{\arraycolsep}{0.5em}
%\left[\begin{array}{ccc}
%0 & 1 & 2 \\
%0 & 1 & 5 \\
%0 & 4 & 2 \\
%0 & 4 & 5 \\
%3 & 1 & 2 \\
%3 & 1 & 5 \\
%3 & 4 & 2 \\
%3 & 4 & 5
%\end{array}
%\right]
\end{equation*}
%
The $2^n$ unique choices for $\bowtie$ on $n$-vectors
corresponds to taking all bit-strings of length $n$ and
selecting from $u$ for $1$ and $v$ for $0$. The pairwise permutation
matrix is thus defined in terms of the logical
matrix $b$ where $b_{i,j}$ is the $j$-th bit of the integer
$i - 1$ as:
%
\begin{align*}
(u \bowtie v)_{i,j} & = b_{i,j} u_j + \neg b_{i,j} v_j
\end{align*}
%
The permutated schemes $x = (u \bowtie v)_i$ added to the model 
$M \otimes N$ are only those which have do not have $\infty$ in 
the constrained dimensions given by each model ($x_j \neq
\infty$). This filters out all permutations which have
$\infty$ in a constrained dimension position. 
\end{definition}

@@

We have explained the model of regions as sets of index
schemes, that is
$\interp{\textit{region}}_n \in \mathcal{P}(\domainVal{}^n)$.
Figure~\ref{fig:spatial-model} defines the top-level model of the
\textit{spec} syntax, which includes the multiplicity and
approximation modifiers. The domain of this interepretation
is $\textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))$
where $\textsf{Mult}$ and $\textsf{Approx}$ are labelled variants
labelling our set-basd model with additional information.

@@

For our example, the \textit{analyse} function matches on
line 5, with the following set for $\textsf{flowsTo}(\texttt{a(i, j) + x +
  y})$:
%
\begin{align*}
\{\texttt{a(i-1, j)}, \texttt{a(i+1, j)}, \texttt{a(i, j-1)},
  \texttt{a(i, j+1)}, \texttt{a(i, j)}\}
\end{align*}
Subsequently the result of \textit{analyse} on line 5 yields the map:
\begin{align*}
\texttt{a} \mapsto \{\vtwoh{-1}{0}, \vtwoh{1}{0},
          \vtwoh{0}{-1}, \vtwoh{0}{1}, \vtwoh{0}{0}\}
\end{align*}
%
If the LHS $\overline{e_1}$ contains non-$0$ neighour offsets 
then all RHS schemes are \emph{relativized} by
$\textsf{schematic}(\overline{e_1})$. For example, 
if \texttt{a(i+1) = b(i)} then relativisation of the RHS by the LHS
produces an analysis which is equivalent to the analysis of \texttt{a(i) = b(i-1)}.

@@

\begin{definition}[Spans]
  A \emph{span} represents an $n$-dimensional box
  (\emph{hyper-rectangle}) as a pair of $n$-dimensional vectors (or
  represented as a $2 \times n$ matrix) with values drawn from
  $\domainVal{}$. This gives the co-ordinates of a lower-bound vertex
  (first comonent) and the upper-bound (second component). We write 
  spans as $\vect{x} = [\vect{x}^L \; \vect{x}^U]$, where $\vect{x}^L$
  and $\vect{x}^U$ give the lower and upper bounds.  In
  the rest, $\vect{x}, \vect{y}$ range over spans.  For a span
  $\vect{x}$, we write $\vect{x}_{i:j}$ for the sub-span comprises the
  subvectors $[\vect{x}^L_{i:j} \; \vect{x}^U_{i:j}]$ from entry $i$
  to entry $j$ (inclusive).
\end{definition}

@@

% taking permustations is a systematic way of producing all rotations
% of the n-dimensional space - forward ref to fold
The next step repeatedly coalesces any contiguous spans
until a fixed point is reached, resulting in a set of (possibly
overlapping) spans covering the original space of schemes. Two $n$-dimensional spans $\vect{x}$ and $\vect{y}$
are contiguous if for all but one $j \in \{1, \ldots, n\}$ then
$\vect{x}_j = \vect{y}_j$ and for one $j$ then $\vect{x}^U_1 + 1 =
\vect{y}^U$. For example, spans
$\left[\vtwoh{-1}{0}\; \vtwoh{-1}{0}\right]$ and $[\vtwoh{0}{0} \; \vtwoh{1}{0}]$ 
are contiguous in the first dimension and can be coalesced
into the span $[\vtwoh{-1}{0} \; \vtwoh{1}{0}]$.

>>>>

\begin{abstract}
  Verifying the correctness of numerical computations relies on first
  giving a specification of their behaviour. In many cases, a
  numerical computation is derived from an underlying mathematical
  model, but this is often difficult to use as a specification as it
  is highly decoupled from the implementation. We have been studying
  specification techniques in the middle ground between low-level and
  high-level specifications for numerical code. In this paper, we
  focus on \emph{stencil computations}---a common idiom in numerical
  code, but one which is error-prone due to fine-grained indexing
  errors. We hypothesise that in practice, stencil computations tend
  to have a regular shape which is amenable to static analysis and
  simple specification. We describe a abstract spatial specification
  language for stencil computations that can be embedded as
  annotations in source code, along with inference, checking, and
  specification synthesis procedures. We evaluate our
  language against a corpus of numerical Fortran code, and show that
  the vast majority of stencil computations indeed have a simple,
  regular, static shape. Around 10\% of the stencils found evoke a
  programming pattern for which indexing errors are plausible. We
  investigate two stencils in detail and find that we are able to
  detect the vast majority of potential errors. This paper details our
  stencil specification language, its static semantics, inference,
  checking, and synthesis procedures, our implementation, and
  our evaluation study.
\end{abstract}

@@

\item we provide an implementation of our approach as an extension to
  CamFort, an open-source program analysis tool for Fortran programs.
  We use this for a quantitative evaluation of our approach over a corpus of
  scientific applications, totalling roughly one million lines of
  code;

@@

In recent years, there has been an increased awareness
of the need for formal verification techniques in
science~\cite{post2005computational,oberkampf2010verification,orchard2014computational}.
Our specification language is an approach in this direction, providing
a system of lightweight specification for numerical code.
The aim is to not only provide a way of exposing errors, but also
for specifying program behaviour to future-proof the code and ease
its maintenance. In order to increase the likelihood of impact, we aim
to integrate with existing practices smoothly, hence our
specifications are integrated as comments and our tool can synthesis
specifications for the user.

@@

\begin{definition}[Neighbourhood index]
  For an array subscript $v(\bar{e})$, we say that $e \in \bar{e}$
  is a \emph{neighbourhood index} if it is of the form
  $e \equiv i$, $e \equiv i + a$ or $e \equiv i - a$ where $a$ is an
  integer constant. That is, a neighbourhood index is a
  relative index defined as constant translations of base induction
  variables. (The relation $\equiv$ here identifies terms up-to commutativity
  of $+$ and the inverse relation of $+$ and $-$ \eg{},
  $(-b) + i \equiv i - b$).  We classify neighbourhood indices using
  the predicate \neigh{}. %An array subscript $v(\bar{e})$ where every
% $e \in \bar{e}$ is a neighbourhood index is called a
% \emph{neighbourhood subscript}.
\label{def:neighbour}
\end{definition}

@@

Figure~\ref{fig:equations} lists the equational theory for
specifications which is defined by the binary relation $\equiv$
between region terms. The equations use abbreviations
\term{refl}, \term{fwd}, \term{bwd}, \term{cen}, and \irreflS{}
for \term{reflexive}, \term{forward}, \term{backward},
\term{centered}, and \irrefl{} respectively. Furthermore, the abbreviations include syntactic sugar
for quantifying over the absence or presence of the \irrefl{}
attribute on region constants as a boolean. For example,
the syntactic sugar for \term{fwd} is defined:
%%
\begin{align*}
\hspace{-0.25em}
& \stenFwdSR{n}{d}{r}
\; := \; \\[-0.4em]
& \qquad \begin{cases}
\stenFwd{n}{d}{, \irrefl} & \;\;\textit{if} \;\, \neg \, r \\
\stenFwd{n}{d}{}  & \;\;\textit{if} \;\,  r
\end{cases} \\
\textit{or} \;\;
& \stenFwdS{n}{d} := \stenFwd{n}{d}{}
\end{align*}
%
and similarly for \term{bwd} and \term{cen}.

@@

\paragraph{Overlapping regions with +}
The first nine
equations (with labels of the form $\ldots \texttt{+} \ldots$)
explain the notion of \emph{region overlapping} with sum
\term{+}. In the first six, $r$ and $s$ range
over booleans denoting the absence (true) or presence (false)
of \irrefl{}. These are combined via disjunction. For example,
the rule $(\textsc{B} \, \term{+} \,
\textsc{F})$ explains that the sum of \term{backward}
and \term{forward} regions in the same dimension and the same
depth is equivalent to a \term{centered} region of that depth.
If the \term{forward} part is irreflexive ($r =
\textit{false}$) but \term{backward} is not ($s =
\textit{true}$) then their sum is equal to a \term{centered}
region with no \irreflS{} modifier by the disjunction ($r \vee s =
\textit{true}$), \ie{}:
%%
\begin{align*}
& \stenCenS{n}{d}  \\[-0.2em]
\equiv \;\; & \stenFwdSR{n}{d}{\irreflS} \; \texttt{+} \; \stenBwdS{n}{d}
\end{align*}
%%
That is, the \term{fwd} region is irreflexive describing code which is missing
a zero-offset neighbourhood index (\eg{}, missing $\texttt{a(i)}$), but the addition of
the \term{bwd} has no irreflexive attribute and thus provides the
missing zero offset index.

@@

Figure~\ref{fig:inequations} defines a notion of sub-specifications
via the relation $<:^r$ on \textit{region}
syntax and then $<:$ on \textit{spec} syntax inductively.
Both relations are reflexive and transitive (we omit
these equations) and congruent
with respect to \term{+}, \term{*}, \term{atLeast}, \term{atMost},
and \term{readOnce}. The (\trule{\textsc{eq}}) rule connects region
equations with inequations. The (\trule{\textsc{over}}) rule explains
the notion of spatial over-approximation via overlapping regions.
The (\trule{\textsc{rep}}) allows the modifier that
enforces that every index is read just once to be dropped
as an approximation; (\trule{\textsc{bound}}) gives
a connection between the \term{atLeast} and \term{atMost}
approximation modifiers.

@@

Indexing schemes give an abstract
representation of index expressions in the source language.
Our model represents the meaning of stencil specifications
as indexing schemes, which explain the range of possible
indexing behaviours allowed under the specification.
The relationship between indexing schemes and source language
terms is given precisely by the partial function, \textsf{schematic}, which
maps syntactic array-index terms to indexing schemes. This
operation is used by the inference
and the checking procedures.
\end{definition}
%
\begin{definition}%[Schematic]
\label{def:schematic}
The partial function $\textsf{schematic}$ maps
source indexing terms $(e_1, .., e_n)$
to $n$-dimensional index schemes, defined:
%
\begin{align*}
\mathsf{schematic}(e_1, \ldots, e_n) & =
[\downarrow\!e_1 \ldots \downarrow\!e_n ]
\quad \textit{iff} \;\; \forall i . \downarrow\!e_i \neq \bot
\end{align*}
%
The $\downarrow$ operator is partial
 $\downarrow : \textit{expr} \rightarrow (\domainVal + \bot)$
mapping individual indexing expressions to 
an indexing scheme component, defined:
%%
\begin{align*}
\downarrow\!e
 & =  \begin{cases}
a & e \equiv i + a \; \wedge \; \mathsf{neigh}(i + a) \\
-a & e \equiv i - a \; \wedge \; \mathsf{neigh}(i - a) \\
\infty & \textit{$\mathsf{IV}(e) = \emptyset$} \\
\bot   & \textit{otherwise}
\end{cases}
\end{align*}
%%
where $\mathsf{IV}(e)$ is the set of free induction variables
in an expression, and $\neigh{}$ classifies expressions which
are constant offsets from an induction variable
(Defn.~\ref{def:neighbour}). In the third case,
if the expression is not a neighbourhood index but is not
a relative index either (\ie{}, not defined in terms
of an induction variable) then it is mapped to $\infty$, otherwise
$\downarrow\!e$ is undefined (returns $\bot$). The $\mathsf{schematic}$ function
is undefined if $\downarrow$ of any of its components $e_i$ is
undefined.

@@

The model is given by the interpretation function $\interp{-}_n$
(where $n$ is the maximum dimensionality of the specification being
modelled) mapping closed\footnote{That is, we assume that there are no
  occurrences of \textit{rvar} in a specification being modelled.  Any
  \emph{open} specification containing region variables can be made
  closed by straightforward syntactic substitution with a (closed)
  \textit{region}.} specifications to sets of
$n$-dimensional indexing schemes. The interpretation is overloaded on
\emph{regions} in Figure~\ref{fig:region-model} and on the top-level
of a specification \textit{spec} in Figure~\ref{fig:spatial-model}.

@@

\begin{example}
For a $2$-dimensional stencil computation, then
\begin{align*}
& \interp{\stenCenS{2}{1}}_2 \\[-0.1em]
& \qquad = \; \{i\singleEntry{1}{2} \,|\, i \in \{\text{-}2,..., 2\}\!\setminus\!0 \}
  \cup  \{ \zeroEntry{1}{2} \,|\, \textit{true} \} \\[-0.2em]
& \qquad = \; \{-2\singleEntry{1}{2}, \, -\singleEntry{1}{2}, \,
  \singleEntry{1}{2}, \, 2\singleEntry{1}{2}\} \cup
  \{\zeroEntry{1}{2}\} \\[-0.2em]
& \qquad = \; \{\vtwoh{-2}{\infty}, \vtwoh{-1}{\infty}, \vtwoh{0}{\infty},
\vtwoh{1}{\infty}, \vtwoh{2}{\infty}\}
\end{align*}
Note the absorbing behaviour of $\infty$ with respect to
multiplication.
\end{example}
\noindent
The last two equations of Fig.~\ref{fig:region-model} give the
models of $\term{+}$ and $\term{*}$. For $\term{+}$ this is the 
straightforward union of the models of subterms. The model of
$\term{*}$ takes the tensor $\otimes$ of subterm models. The tensor is
non-trivial; we give some informal intuition first.

@@

The intuition behind $\otimes$ is that it takes
all possible pairs of indexing schemes, treating them as the
lower and upper bounds of an $n$-dimensional rectangle from which the
remaining vertices of the rectangle are generated (like a bounding
box) \eg{}:
%%
\begin{align*}
\{\vtwohs{1}{2}, \vtwohs{5}{6}\,\} \otimes \{\vtwohs{3}{4}\} =
\{\vtwohs{1}{2}, \vtwohs{3}{2}, \vtwohs{1}{4}, \vtwohs{3}{4}, 
   \vtwohs{3}{6}, \vtwohs{5}{4}, \vtwohs{5}{6}\,\}
\end{align*}
%%
The tensor takes a Cartesian product of the two models,
giving all pairs of schemes which are combined using the
\emph{pairwise permutation} written $\bowtie$ (defined later).
Pairwise permutation generates all possible interleavings of two vectors
(\ie{}, all possible non-deterministic choices between $u_i$ and
$v_i$ in each dimension). Some care is then taken over
how this interacts with $\infty$: pairwise permutation
corresponds to non-deterministic choice, in each dimension $i$,
between two values $u_i$ and $v_i$, but if either
is $\infty$ then the resulting pairwise permutation is
filtered out if the dimension $i$ has \emph{any other scheme
in the model where $u_i \neq \infty$}, \eg{}:
\begin{align*}
\{\vtwoh{1}{\infty}\} \otimes \{\vtwoh{3}{4}\} =
\{\vtwoh{1}{4}, \vtwoh{3}{4}\}
\end{align*}
%
where $\vtwoh{1}{\infty}$ and $\vtwoh{3}{\infty}$ have been filtered.
%With respect to the notion of unification, the
%
%
% combines non-$\infty$ constraints
%of the unifiers in different dimensions into less accepting unifiers.
%This means an access pattern generalises with this scheme would require
%both models to be satisfied. Use of $dims$ gives the dimensions that
%constituent models talk about and is used in the product to eliminate
%general unifier appearing in these dimensions. Allowing general unifier
%in these dimensions would prevent both models to be satisfied
%simultaneously.
%  u in M * N   u < w
% implies that
%  u_1 in M  u1 < w
%  u_2 in M  u2 < w
%
The notion of which dimensions $i$ have that \emph{any scheme in the model
has $u_i \neq \infty$} is represented by the \emph{constrained
dimensions} intermediate function:
%
\begin{definition}%[Constrained dimensions]
For an $n$-dimensional model $M$, its \emph{constrained dimensions}
is the set of dimension identifiers for which at least one scheme $u \in M$ has
a non-$\infty$ value in that dimension. That is:
%
\begin{align*}
\effdims{M}{n} =
\bigcup_{u \in M} \{i \mid i \in \{1,\ldots,n\} \, \wedge \, u_i \neq
  \infty\}
\end{align*}
For example, $\effdims{\{\vtwoh{0}{\infty},\vtwoh{\infty}{\infty}\}}{2}
 = \{1\}$.
\end{definition}

@@

The \emph{pairwise permutation} $u \bowtie v$ 
builds a $2^n \times n$ matrix 
of all possible $n$-vectors generated by
non-deterministically picking for each $j^{th}$ entry either 
$u_j$ or $v_j$ \eg{}
%
\begin{equation*}
\vthreeh{0}{1}{2} \bowtie \vthreeh{3}{4}{5} =
\left[
\setlength{\arraycolsep}{0.42em}
\begin{array}{cccccccc}
0 & 0 & 0 & 0 & 3 & 3 & 3 & 3 \\
1 & 1 & 4 & 4 & 1 & 1 & 4 & 4 \\
2 & 5 & 2 & 5 & 2 & 5 & 2 & 5
\end{array}
\right]
% \\
%\setlength{\arraycolsep}{0.5em}
%\left[\begin{array}{ccc}
%0 & 1 & 2 \\
%0 & 1 & 5 \\
%0 & 4 & 2 \\
%0 & 4 & 5 \\
%3 & 1 & 2 \\
%3 & 1 & 5 \\
%3 & 4 & 2 \\
%3 & 4 & 5
%\end{array}
%\right]
\end{equation*}
%
The $2^n$ unique choices for $\bowtie$ on $n$-vectors
corresponds to taking all bit-strings of length $n$ and
selecting from $u$ for $1$ and $v$ for $0$. The pairwise permutation
matrix is thus defined in terms of the logical
matrix $b$ where $b_{i,j}$ is the $j$-th bit of the integer
$i - 1$ as:
%
\begin{align*}
(u \bowtie v)_{i,j} & = b_{i,j} u_j + \neg b_{i,j} v_j
\end{align*}
%
The permuted schemes $x = (u \bowtie v)_i$ added to the model 
$M \otimes N$ are only those which have do not have $\infty$ in 
the constrained dimensions given by each model ($x_j \neq
\infty$). This filters out all permutations which have
$\infty$ in a constrained dimension position. 
\end{definition}

@@

We have explained the model of regions as sets of index
schemes, that is
$\interp{\textit{region}}_n \in \mathcal{P}(\domainVal{}^n)$.
Figure~\ref{fig:spatial-model} defines the top-level model of the
\textit{spec} syntax, which includes the multiplicity and
approximation modifiers. The domain of this interpretation
is $\textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))$
where $\textsf{Mult}$ and $\textsf{Approx}$ are labelled variants
labelling our set-based model with additional information.

@@

For our example, the \textit{analyse} function matches on
line 5, with the following set for $\textsf{flowsTo}(\texttt{a(i, j) + x +
  y})$:
%
\begin{align*}
\{\texttt{a(i-1, j)}, \texttt{a(i+1, j)}, \texttt{a(i, j-1)},
  \texttt{a(i, j+1)}, \texttt{a(i, j)}\}
\end{align*}
Subsequently the result of \textit{analyse} on line 5 yields the map:
\begin{align*}
\texttt{a} \mapsto \{\vtwoh{-1}{0}, \vtwoh{1}{0},
          \vtwoh{0}{-1}, \vtwoh{0}{1}, \vtwoh{0}{0}\}
\end{align*}
%
If the LHS $\overline{e_1}$ contains non-$0$ neighbour offsets 
then all RHS schemes are \emph{relativised} by
$\textsf{schematic}(\overline{e_1})$. For example, 
if \texttt{a(i+1) = b(i)} then relativisation of the RHS by the LHS
produces an analysis which is equivalent to the analysis of \texttt{a(i) = b(i-1)}.

@@

\begin{definition}[Spans]
  A \emph{span} represents an $n$-dimensional box
  (\emph{hyper-rectangle}) as a pair of $n$-dimensional vectors (or
  represented as a $2 \times n$ matrix) with values drawn from
  $\domainVal{}$. This gives the co-ordinates of a lower-bound vertex
  (first component) and the upper-bound (second component). We write 
  spans as $\vect{x} = [\vect{x}^L \; \vect{x}^U]$, where $\vect{x}^L$
  and $\vect{x}^U$ give the lower and upper bounds.  In
  the rest, $\vect{x}, \vect{y}$ range over spans.  For a span
  $\vect{x}$, we write $\vect{x}_{i:j}$ for the sub-span comprises the
  subvectors $[\vect{x}^L_{i:j} \; \vect{x}^U_{i:j}]$ from entry $i$
  to entry $j$ (inclusive).
\end{definition}

@@

% taking permutations is a systematic way of producing all rotations
% of the n-dimensional space - forward ref to fold
The next step repeatedly coalesces any contiguous spans
until a fixed point is reached, resulting in a set of (possibly
overlapping) spans covering the original space of schemes. Two $n$-dimensional spans $\vect{x}$ and $\vect{y}$
are contiguous if for all but one $j \in \{1, \ldots, n\}$ then
$\vect{x}_j = \vect{y}_j$ and for one $j$ then $\vect{x}^U_1 + 1 =
\vect{y}^U$. For example, spans
$\left[\vtwoh{-1}{0}\; \vtwoh{-1}{0}\right]$ and $[\vtwoh{0}{0} \; \vtwoh{1}{0}]$ 
are contiguous in the first dimension and can be coalesced
into the span $[\vtwoh{-1}{0} \; \vtwoh{1}{0}]$.
